Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-04-02 paper/manuscript.Rmd
add wg acknowledge
small text edits

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. Only `r n10` of these 12 individuals had more than 10
swimming events, suggesting that for ~10% of collared individuals swim
relatively frequently, on average occurring approximately every `r indMean` days
during ice-free seasons. We did not detect any swimming events for the remaining
17 individuals. It remains unclear whether these individuals do not swim, or
whether swimming is rare, and thus undetected, for the majority of individuals
Overall, there appears to be three groups of individuals: those that swim often
(n = 3, every ~`r indMean` days), those that swim occasionally (n = 9, every ~
`r ind2Mean` days), and those that never swam (n = 17). We are also unable to draw
conclusions about the prevalence of swimming in the ocean by male caribou: our
initial visual observation was of an adult male, but our GPS data only included
females.

@@

For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached ~300 animals and it is unknown when caribou began to occupy other
islands, although anecdotal evidence suggests it was during the 1990s. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation, our data suggest caribou only remain on smaller islands for a few
days at a time, so the fitness equilibrium remains a theoretical construct. More
likely, however, is that fine-scale competition, density-dependent habitat
selection, and forage depletion drives individuals to peridiodically swim
between islands. Given the presumed low density of coyotes on Fogo Island and
the minimal risk they pose, predation risk does not appear to be a major driver
of swimming.

@@

**Acknowledgements**

We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided by L. Bixby. We also thank members of the
Newfoundland Wildlife Division including S. Moores, B. Adams, C. Doucet, W.
Barney, and J. Neville for logistical support in the field as well as help with
data data collection and management. We thank T. Bergerud and S. Mahoney for
their vision in initiating much of the work on caribou in Newfoundland. Funding
for this study was provided by the National Sciences and Engineering Research
Council (QMRW, JGH, EVW). We respectfully acknowledge the territory in which
data were collected and analyzed as the ancestral homelands of the Beothuk, and
the island of Newfoundland as the ancestral homelands of the Mi’kmaq and
Beothuk.
<!-- Simplified the funding statement, do we need to name the specific
scholarships/grants? Eric opinion??-->

>>>>

In total, 12 of 29 collared female caribou swam between islands in the Fogo
Island archipelago. Only `r n10` of these 12 individuals had more than 10
swimming events, suggesting that for ~10% of collared individuals swim
relatively frequently, on average occurring approximately every `r indMean` days
during ice-free seasons. We did not detect any swimming events for the remaining
17 individuals. It remains unclear whether these individuals do not swim, or
whether swimming is rare, and thus undetected, for the majority of individuals
Overall, there appears to be three groups of individuals: those that swim often
(n = 3, every ~`r indMean` days), those that swim occasionally (n = 9, every ~
`r ind2Mean` days), and those that never swam (n = 17). We are also unable to draw
conclusions about the prevalence of swimming in the ocean by male caribou: our
initial visual observation was of an adult male, but our GPS data only included adult
females.

@@

For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached ~300 animals and it is unknown when caribou began to occupy other
islands, although anecdotal evidence suggests it was during the 1990s. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation, our data suggest caribou only remain on smaller islands for a few
days at a time, so the fitness equilibrium remains a theoretical construct. More
likely, however, is that fine-scale competition, density-dependent habitat
selection, and forage depletion drives individuals to periodically swim
between islands. Given the presumed low density of coyotes on Fogo Island and
the minimal risk they pose, predation risk does not appear to be a major driver
of swimming.

@@

**Acknowledgements**
We thank M. Laforge, M. Bonar, C. Hart, and R. Huang for help in the field.
Logistical support was provided by L. Bixby. We thank all members of the
Wildlife Evolutionary Ecology Lab, including C. Hart, C. Prokopenko, J. Kennah,
J.W. Turner, and S. Boyle for their comments on previous versions of this
manuscript. Newfoundland Wildlife Division including S. Moores, B. Adams, C.
Doucet, W. Barney, and J. Neville for logistical support in the field as well as
help with data data collection and management. We thank T. Bergerud and S.
Mahoney for their vision in initiating much of the work on caribou in
Newfoundland. Funding for this study was provided by the National Sciences and
Engineering Research Council (QMRW, JGH, EVW). We respectfully acknowledge the
territory in which data were collected and analyzed as the ancestral homelands
of the Beothuk, and the island of Newfoundland as the ancestral homelands of the
Mi’kmaq and Beothuk.
<!-- Simplified the funding statement, do we need to name the specific
scholarships/grants? Eric opinion??-->
